Modeling Intervention Strategies for United States TB Control

### 2014-02-08 in_progress/presentations/paper/draft.tex
Modifying discussion to discuss plateauing effect

Further implementations of the \texttt{C++} model were made that deviated from
the basic Hill Model. Specifically, the acute latent compartment in the basic
Hill Model is a vestigal necessity of compartmental modeling and not reflective
of the biology of tuberculosis, and an agent based model was implemented that
more closely respects the biology of tuberculosis. Results from this model or
the strict Hill model did not differ significantly. 
\section{Results}
\subsection{Basic Population Breakdown}
The additional tracking capabilities offered several key insights into US TB
dynamics. Figure~\ref{fig:incPlotSourced} shows the yearly incidence of US TB, broken
down into infection source. It can be seen that the majority of the US TB load
is driven by activations of FB LTBI, followed by USB LTBI activations. This data
further agrees with the conclusions drawn by Hill, Becerra, and Castro about the
necessity of LTBI treatment in any valuable intervention strategy. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.5]{incPlotSourced}
  \end{center}
  \caption{Sourced yearly incidence data generated by the extended Hill Model.}
  \label{fig:incPlotSourced}
\end{figure}
Further, figure~\ref{fig:costPlotSourced} shows a similarly sourced plot, but analyzing the
final US HCS costs due to TB. One can see that in this plot, roughly half of the
US TB HCS costs are due to activations of LTBI. In this data, it is appropriate
to think of the costs due to acute LTBI are due to exogenous re-inection,
whereas active TB costs due to chronic LTBI are activations of LTBI into active
TB. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.35]{costPlotSourced}
  \end{center}
  \caption{Sourced US HCS economic TB load. Note that this data only illustrates
    the load due to treating active TB, but illustrates where the infections
    driving this cost come from.}
  \label{fig:costPlotSourced}
\end{figure}
Note that both of these plots underestimate the impact LTBI activations play in
the spread of TB, as every LTBI activation to infectious TB contributes not only
to incidence and costs directly, but also indirectly by causing additional
future cases, which is not captured in these graphs. Further, there is also US
HCS costs due to treating LTBI, which is not illustrated in these graphs. 
\subsubsection{Basic Reproduction Number}
The basic reproduction number of FB or USB cases of infectious TB was also
estimated by this system. Using the best-fit values of parameters from the 
Hill model, we found the average number of secondary infections arising from
a single case of infectious TB to be 0.423 and 0.370 for USB and FB populations
respectively.  These values differ because in the model, there are different
estimates for death rates and the rate of TB treatment and self-cure in the USB and FB populations, 
so the average infectious period differs between the two populations.
Extrapolating these results, we can think of the
total number of secondary infections over 100 years due to a FB or USB
infectious TB case as describing a geometric series in a large population.
Presuming there are no overlaps in infectious contacts, if a single case of
infectious TB in either population infects $p_f, p_u$ new cases,
respectively, then we can say that over the course of 100 years, the total
number of cases infected will follow a geometric series. This analysis predicts
that over 100 years, one USB infection will lead to $1.04$ subsequent infections,
whereas one FB infection will lead to $.83$ subsequent infections. Experimentally, 
these data were also analyzed, with results of $1.03$ and $.64$, respectively. 
Full calculations are included in the Appendix.

@@

% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Oct 14 00:36:15 2013
\begin{table}
\centering
\begin{tabular}{|r|cccc|}
  \hline
 & red5 & red10 & red25 & red50 \\ 
  \hline
2000 & ND & ND & ND & ND \\ 
  2025 & 67908.81 & 67855.46 & 67654.07 & 67332.53 \\ 
  2050 & 28938.23 & 28878.70 & 28699.15 & 28404.13 \\ 
  2075 & 21044.88 & 20989.16 & 20820.53 & 20541.77 \\ 
  2100 & 18130.11 & 18076.11 & 17912.95 & 17642.64 \\ 
   \hline
\end{tabular}
\caption{Cost Per Case Averted by Reducing Incoming LTBI by X percent (in dollar per case)} 
\label{tab:cpcaArt}
\end{table}

@@

\subsection{Sensitivity Analysis}
To account for uncertainty of input parameter values and to gain insight about the
most influential parameters in the model, we used Latin Hypercube Sampling to analyze the deterministic model, implemented with the \texttt{lhs} package in \texttt{R}.  We varied 
the 16 parameters from the original Hill model according to triangular distributions centered
around their best-fit values, and two additional treatment cost parameters according to uniform distributions.  We obtained similar results to the Hill model 
analyzing non-economic outcomes, as expected for validation.  For economic 
outcomes, we found the parameter $f$, the fraction of FB arrivals with LTBI,
to be highly correlated with the projected overall cost burden of TB in the United States
over the next 100 years.  Another highly influential parameter in the model was $\sigma^{L}$, 
the treatment rate for chronic LTBI.  However, while increasing $\sigma^{L}$ significantly
decreases the cost burden due to Active TB treatment, it increases the cost burden due to Latent TB treatment by a greater amount, given the model estimates that Latent TB treament and Active
TB treatment health care costs are approximately \$700 and \$14,000 per case cured respectively.  
A full description of the Latin Hypercube Sampling analysis and tables of Partial Rank Correlation Coefficient (PRCC) results for all parameters are given in the Appendix, Section 7.6.  \\

@@

\section{Discussion}
These results confirm the hypothesis that curing incoming LTBI rates is a
necessary step towards elimination and indicate that it is a cost effective
option. \\
{\huge \color{red} Where did we show it was cost effective relative to other
interventions? Also, the section below (future work) might be better just
directly included in the discussion}\\
\subsection{Future Work}
This work could be extended by examining different classes of interventions or
more accurately estimating intervention cost with the deterministic extended
Hill. Further work could also be done with the agent-based Hill Model, by using
it to examine the effect contact structure plays on US TB incidence levels or to
examine the effects drug-resistant TB will have on US TB dynamics.

@@

% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Oct 14 00:36:15 2013
\begin{table}
\centering
\begin{tabular}{|r|cccc|} \hline
       & red5     & red10    & red25    & red50    \\ \hline
  2000 & ND       & ND       & ND       & ND       \\ 
  2025 & 67908.81 & 67855.46 & 67654.07 & 67332.53 \\ 
  2050 & 28938.23 & 28878.70 & 28699.15 & 28404.13 \\ 
  2075 & 21044.88 & 20989.16 & 20820.53 & 20541.77 \\ 
  2100 & 18130.11 & 18076.11 & 17912.95 & 17642.64 \\ \hline
\end{tabular}
\caption{Cost Per Case Averted by Reducing Incoming LTBI by various percent (in
dollars per case, Cure=\$800)} 
\label{tab:cpcaRed}
\end{table}

>>>>

Further implementations of the \texttt{C++} model were made that deviated from
the basic Hill Model. Specifically, the acute latent compartment in the basic
Hill Model is a vestigal necessity of compartmental modeling and not reflective
of the biology of tuberculosis, and an agent based model was implemented that
more closely respects the biology of tuberculosis. Results from this model or
the strict Hill model did not differ significantly.

\section{Results}
\subsection{Basic Population Breakdown}
The additional tracking capabilities offered several key insights into US TB
dynamics. Figure~\ref{fig:incPlotSourced} shows the yearly incidence of US TB, broken
down into infection source. It can be seen that the majority of the US TB load
is driven by activations of FB LTBI, followed by USB LTBI activations. This data
further agrees with the conclusions drawn by Hill, Becerra, and Castro about the
necessity of LTBI treatment in any valuable intervention strategy. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.5]{incPlotSourced}
  \end{center}
  \caption{Sourced yearly incidence data generated by the extended Hill Model.}
  \label{fig:incPlotSourced}
\end{figure}
Further, figure~\ref{fig:costPlotSourced} shows a similarly sourced plot, but analyzing the
final US HCS costs due to TB. One can see that in this plot, roughly half of the
US TB HCS costs are due to activations of LTBI. In this data, it is appropriate
to think of the costs due to acute LTBI are due to exogenous re-inection,
whereas active TB costs due to chronic LTBI are activations of LTBI into active
TB. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.35]{costPlotSourced}
  \end{center}
  \caption{Sourced US HCS economic TB load. Note that this data only illustrates
    the load due to treating active TB, but illustrates where the infections
    driving this cost come from.}
  \label{fig:costPlotSourced}
\end{figure}
Note that both of these plots underestimate the impact LTBI activations play in
the spread of TB, as every LTBI activation to infectious TB contributes not only
to incidence and costs directly, but also indirectly by causing additional
future cases, which is not captured in these graphs. Further, there is also US
HCS costs due to treating LTBI, which is not illustrated in these graphs. 
\subsubsection{Basic Reproduction Number}
The basic reproduction number of FB or USB cases of infectious TB was also
estimated by this system. Using the best-fit values of parameters from the 
Hill model, we found the average number of secondary infections arising from
a single case of infectious TB to be 0.423 and 0.370 for USB and FB populations
respectively.  These values differ because in the model, there are different
estimates for death rates and the rate of TB treatment and self-cure in the USB and FB populations, 
so the average infectious period differs between the two populations.
Extrapolating these results, we can think of the
total number of secondary infections over 100 years due to a FB or USB
infectious TB case as describing a geometric series in a large population.
Presuming there are no overlaps in infectious contacts, if a single case of
infectious TB in either population infects $p_f, p_u$ new cases,
respectively, then we can say that over the course of 100 years, the total
number of cases infected will follow a geometric series. This analysis predicts
that over 100 years, one USB infection will lead to $1.04$ subsequent infections,
whereas one FB infection will lead to $.83$ subsequent infections. Experimentally, 
these data were also analyzed, with results of $1.03$ and $.64$, respectively. 
Full calculations are included in the Appendix.

@@

% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Oct 14 00:36:15 2013
\begin{table}
\centering
\begin{tabular}{|r|cccc|} \hline
       & red5     & red10    & red25    & red50    \\ \hline
  %2000 & ND       & ND       & ND       & ND       \\ 
  2025 & 67908.81 & 67855.46 & 67654.07 & 67332.53 \\ 
  2050 & 28938.23 & 28878.70 & 28699.15 & 28404.13 \\ 
  2075 & 21044.88 & 20989.16 & 20820.53 & 20541.77 \\ 
  2100 & 18130.11 & 18076.11 & 17912.95 & 17642.64 \\ \hline
\end{tabular}
\caption{Cost Per Case Averted by Reducing Incoming LTBI by X percent (in dollar per case)} 
\label{tab:cpcaArt}
\end{table}

@@

\subsection{Sensitivity Analysis}
To account for uncertainty of input parameter values and to gain insight about
the most influential parameters in the model, we used Latin Hypercube Sampling
to analyze the deterministic model, implemented with the \texttt{lhs} package in
\texttt{R}.  We varied the 16 parameters from the original Hill model according
to triangular distributions centered around their best-fit values, and two
additional treatment cost parameters according to uniform distributions.  We
obtained similar results to the Hill model analyzing non-economic outcomes, as
expected for validation.  For economic outcomes, we found the parameter $f$, the
fraction of FB arrivals with LTBI, to be highly correlated with the projected
overall cost burden of TB in the United States over the next 100 years.  Another
highly influential parameter in the model was $\sigma^{L}$, the treatment rate
for chronic LTBI.  However, while increasing $\sigma^{L}$ significantly
decreases the cost burden due to Active TB treatment, it increases the cost
burden due to Latent TB treatment by a greater amount, given the model estimates
that Latent TB treament and Active TB treatment health care costs are
approximately \$700 and \$14,000 per case cured respectively.  A full
description of the Latin Hypercube Sampling analysis and tables of Partial Rank
Correlation Coefficient (PRCC) results for all parameters are given in the
Appendix, Section 7.6.

@@

\section{Discussion}
These results confirm the hypothesis that curing incoming LTBI rates is a
necessary step towards elimination and indicate that it is a cost effective
option. \\
{\huge \color{red} Where did we show it was cost effective relative to other
interventions? Also, the section below (future work) might be better just
directly included in the discussion. Also, if we want to make this claim here we
need to include other interventions in the main body.}\\
In addition, beyond demonstrating quantitatively that this intervention strategy
is a necessary step towards US elimination, we also obtained qualitative
indications that this is the case. In every intervention we tested (see, for
example, Intervention <+ADD ME+> and <+ADD ME+> in section <+ADD ME+>) we
observed a plateauing effect, where the incidence per million stabilized after
some time and resisted further change. In particular, even when transmission was
reduced to $0\%$, we still observed this plateauing effect, which implies that
something other than active transmission was generating enough incidence to
sustain TB in an endemic state in the US population. However, when the FB LTBI
immigration was cut, this effect vanished. These results therefore support the
argument that the USB LTBI population is sufficiently large to sustain a
sizable, endemic TB-disease state in the United States provided it is
continually refueled via the entering LTBI immigrants. Thus, there is no
intervention that can ignore the LTBI population that can hope to eliminate TB
in the United States. Among those interventions that treat cases of LTBI,
targeting the immigrating LTBI requires no additional US HCS dollars beyond what
is spent now to identify the LTBI population and is therefore extremely
efficient. Additionally, our analysis shows that paying for the treatment of
these individuals is extremely cost effective {\huge \color{red} TODO: Show
this}.
\subsection{Future Work}
This work could be extended by examining different classes of interventions or
more accurately estimating intervention cost with the deterministic extended
Hill. Further work could also be done with the agent-based Hill Model, by using
it to examine the effect contact structure plays on US TB incidence levels or to
examine the effects drug-resistant TB will have on US TB dynamics.

@@

% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Oct 14 00:36:15 2013
\begin{table}
\centering
\begin{tabular}{|r|cccc|} \hline
       & red5     & red10    & red25    & red50    \\ \hline
  %2000 & ND       & ND       & ND       & ND       \\ 
  2025 & 67908.81 & 67855.46 & 67654.07 & 67332.53 \\ 
  2050 & 28938.23 & 28878.70 & 28699.15 & 28404.13 \\ 
  2075 & 21044.88 & 20989.16 & 20820.53 & 20541.77 \\ 
  2100 & 18130.11 & 18076.11 & 17912.95 & 17642.64 \\ \hline
\end{tabular}
\caption{Cost Per Case Averted by Reducing Incoming LTBI by various percent (in
dollars per case, Cure=\$800)} 
\label{tab:cpcaRed}
\end{table}
